Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-06 treestats_paper.tex
Remove Indeed,

\krt{Why? I think this is an odd way of putting it.  The challenge back then was
    genotyping.  Once that was done, summarizing the relationships was done on a 
    calculator b/c the data sets were so small.}
It was once a major undertaking to estimate a single summary
of the many genetic relationships between the individuals of a sample,
e.g., to estimate $F_{ST}$ between two populations using a dozen allozymes.
Today's vast quantity of whole-genome sequence
makes it possible to confidently estimate many more properties of genealogical relationships
in local regions of genomes \citep{stankowski2019widespread,fst_landscapes},
or between individuals rather than populations \citep{individual_stats}.
Computation is beginning to be a major problem:
whole-genome sequence on the scale of the UK Biobank,
for instance, might have 500K individuals genotyped at 10M loci \citep{ukbb},
resulting in a genotype matrix of $5 \times 10^{12}$ entries.
Computing a single summary from this matrix is still relatively quick,
once loaded into memory,
but for many applications we are now interested in computing a large number of statistics,
or iteratively refining a single computation.
Indeed, many sorts of modern inference machinery, such as deep learning,
depend on being able to compute enough statistics that this becomes a serious bottleneck in practice.

>>>>

\krt{Why? I think this is an odd way of putting it.  The challenge back then was
    genotyping.  Once that was done, summarizing the relationships was done on a 
    calculator b/c the data sets were so small.}
It was once a major undertaking to estimate a single summary
of the many genetic relationships between the individuals of a sample,
e.g., to estimate $F_{ST}$ between two populations using a dozen allozymes.
Today's vast quantity of whole-genome sequence
makes it possible to confidently estimate many more properties of genealogical relationships
in local regions of genomes \citep{stankowski2019widespread,fst_landscapes},
or between individuals rather than populations \citep{individual_stats}.
Computation is beginning to be a major problem:
whole-genome sequence on the scale of the UK Biobank,
for instance, might have 500K individuals genotyped at 10M loci \citep{ukbb},
resulting in a genotype matrix of $5 \times 10^{12}$ entries.
Computing a single summary from this matrix is still relatively quick,
once loaded into memory,
but for many applications we are now interested in computing a large number of statistics,
or iteratively refining a single computation.
Many sorts of modern inference machinery, such as deep learning,
depend on being able to compute enough statistics that this becomes a serious bottleneck in practice.
